---
title: Using the package
output: html_document
---

We now present some of the main features of the package. We notice that the package supports different types of mutational signatures such as: SBS (single base substitutions) and MNV (multi-nucleotide variant) (see Degasperi, Andrea, et al. "Substitution mutational signatures in whole-genome–sequenced cancers in the UK population." Science 376.6591 (2022): abl9283), CX (chromosomal instability) (see Drews, Ruben M., et al. "A pan-cancer compendium of chromosomal instability." Nature 606.7916 (2022): 976-983) and CN (copy number) signatures (see Steele, Christopher D., et al. "Signatures of copy number alterations in human cancer." Nature 606.7916 (2022): 984-991). But, for the sake of this vignette, we present only results on the classical SBS signatures. We refer to the manual for details.

First, we show how to load example data and import them into a count matrix to perform the signatures analysis.

```{r}
library("SparseSignaturesPlus")
data(ssm560_reduced)
```

These data are a reduced version of the 560 breast tumors provided by Nik-Zainal, Serena, et al. (2016) comprising only 3 patients. We notice that these data are provided purely as an example, and, as they are a reduced and partial version of the original dataset, they should not be used to draw any biological conclusion.

We now import such data into a count matrix to perform the signatures discovery. To do so, we also need to specify the reference genome as a BSgenome object to be considered. This can be done as follows, where in the example we used hs37d5 as reference genome as provided as data object within the package.

```{r}
library("BSgenome.Hsapiens.1000genomes.hs37d5")
imported_data = getSBSCounts(data = ssm560_reduced, reference = BSgenome.Hsapiens.1000genomes.hs37d5)
head(imported_data)
```

Now, we present an example of visualization feature provided by the package, showing the counts for the first patient, i.e., PD10010a, in the following plot.

```{r fig.width=7, fig.height=7, fig.cap="Visualization of the counts for patient PD10010a from the dataset published in Nik-Zainal, Serena, et al."}
patientsSBSPlot(trinucleotides_counts=imported_data,samples="PD10010a")
```

After the data are loaded, we can perform signatures de novo extraction. To do so, we need to define a range for the number of signatures (variable K) to be considered. We now show how to perform the inference on the first 3 patients of the dataset from Nik-Zainal, Serena, et al. (2016), whose counts are provided within the package.

```{r}
data(background)
data(patients)
set.seed(12345)
res_denovo = signaturesDecomposition(x = patients[1:3,], 
                                     K = 3:4, 
                                     background_signature = background, 
                                     nmf_runs = 2, 
                                     sparsify = FALSE, 
                                     num_processes = 1)
```

Now that we have performed the de novo inferece, we need to decide the optimal number of signatures to be extracted from our dataset. To do so, we provide a procedure based on cross-validation.

```{r}
set.seed(12345)
res_cv = signaturesCV(x = patients[1:3,], 
                      beta = res_denovo$beta, 
                      cross_validation_iterations = 2, 
                      cross_validation_repetitions = 2, 
                      num_processes = 1)
```

We notice that the computations for this task can be very time consuming, expecially when many iterations of cross validations are performed (see manual) and a large set of configurations of the parameters are tested.

We conclude this vignette by plotting the discovered signatures for the best configuration, i.e., K = 4.

```{r fig.width=7, fig.height=7, fig.cap="Visualization of the discovered signatures."}
signatures = res_denovo$beta[[2]]
signaturesSBSPlot(beta=signatures, xlabels=FALSE)
```

We refer to the manual for a detailed description of each parameter and to the SparseSignaturesPlus manuscript for details on the method.
